A Large-scale Study of Polymorphism in R

### 2019-03-23 notebooks/paper_results_section.Rmd
plots and new data, with vector/scalar dealt with

---
title: "Overview"
output: html_notebook
---

Ok.
First, load the full signatures, and make sure we source the data processing file.
There's an error here, not sure what's causing it.
This is a work-in-progress.

```{r}
source("../code/data_processing.R") # get the functions we need
```

```{r, echo=FALSE}
los <- list.files("../data/full_sigs", full.names=TRUE)
names <- list.files("../data/full_sigs")
names(los) <- substr(names, 1, nchar(names)-4)
los <- lapply(los, readRDS)
los <- remove_empties(los)
# annoying strange error that I should fix pops up
```

These are the updated counts.
```{r}
# argument counts
total_arguments <- apply_count_to_lopkg_c(los, args_count_all)
total_mono_type <- apply_count_to_lopkg_c(los, args_count_all_mono_type)
total_mono_attr <- apply_count_to_lopkg_c(los, args_count_all_mono_attr_pattern)
total_mono_clas <- apply_count_to_lopkg_c(los, args_count_all_mono_class)
total_mono_all <- apply_count_to_lopkg_c(los, args_count_all_mono_all)
```

@@

This is how you get the top sigs.
```{r}
poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
```

Then, pick out the ones you want.
The `format_df_for_print` function is just for letting the notebook display the data frame properly.
The type, attr, and class rows are lists of lists.
```{r}
format_df_for_print(poly_type_args[1:20,])
```

Similarly for attributes, and classes.
```{r}
poly_attr_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_attr_args))
poly_attr_args <- poly_attr_args[order(-poly_attr_args$count),] # sort by count
poly_attr_args$perc <- round(poly_attr_args$count/sum(poly_attr_args$count) * 100, 2)
```

```{r}
poly_class_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_class_args))
poly_class_args <- poly_class_args[order(-poly_class_args$count),] # sort by count
poly_class_args$perc <- round(poly_class_args$count/sum(poly_class_args$count) * 100, 2)
```

Printing is as before.
Note that ATM I also have attribute type information in here.
I'll get it to display just the names as well.
I figured this might be interesting, and I don't expect many functions to be polymorphic in this way.
```{r}
format_df_for_print(poly_attr_args[1:20,])
```

```{r}
format_df_for_print(poly_class_args[1:20,])
```

How many things are just scalar?
```{r}
lo_scalars <- apply_row_extractor_to_lopkg_c(los, get_rows_from_df_only_scalar)
total_scalar_arguments <- apply_count_to_lopkg_c(lo_scalars, args_count_all)
```

Let's see what happens if we change type systems!
```{r}
los_basic <- lapply(los, function(lof) {lapply(lof, function(df) {change_type_systems_df(df, type_map_T0_to_r)})})
```

Let's get some counts.
This is for a type system which is in direct correspondence with R's typeof.
```{r}
# argument counts
total_mono_type_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_type)
total_mono_all_basic <- apply_count_to_lopkg_c(los_basic, args_count_all_mono_all)
```

What if we roll NULL into every type?
```{r}
los_basic_no_null <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_into_other_types_df(df)})})

# also curious if we roll raw_NAs in
los_basic_no_null_no_NA <- lapply(los_basic, function(lof) {lapply(lof, function(df) {fold_NULL_and_NA_into_other_types_df(df)})})
```

What are the counts now?
```{r}
total_mono_type_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_type)
total_mono_all_basic_no_null <- apply_count_to_lopkg_c(los_basic_no_null, args_count_all_mono_all)
```

Great.
Now, what are the top instances of type-based polymorphism?
```{r}
poly_type_args_basic_no_null <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null, extract_poly_type_args))
poly_type_args_basic_no_null <- poly_type_args_basic_no_null[order(-poly_type_args_basic_no_null$count),] # sort by count
poly_type_args_basic_no_null$perc <- round(poly_type_args_basic_no_null$count/sum(poly_type_args_basic_no_null$count) * 100, 2)
```

What's the top?
```{r}
format_df_for_print(poly_type_args_basic_no_null[1:20,])
```

So, clearly, "real number polymorphism" is the top choice.
What if we fold ints and doubles together into reals?

```{r}
los_basic_no_null_real <- lapply(los_basic_no_null, function(lof) {lapply(lof, function(df) {fold_together_int_double_df(df)})})
```

What are the counts now?
```{r}
total_mono_type_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_type)
total_mono_all_basic_no_null_real <- apply_count_to_lopkg_c(los_basic_no_null_real, args_count_all_mono_all)
```

Great.
Now, what are the top instances of type-based polymorphism?
```{r}
poly_type_args_basic_no_null_real <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_basic_no_null_real, extract_poly_type_args))
poly_type_args_basic_no_null_real <- poly_type_args_basic_no_null_real[order(-poly_type_args_basic_no_null_real$count),] # sort by count
poly_type_args_basic_no_null_real$perc <- round(poly_type_args_basic_no_null_real$count/sum(poly_type_args_basic_no_null_real$count) * 100, 2)
```

What's the top?
```{r}
format_df_for_print(poly_type_args_basic_no_null_real[1:20,])
```

Ok.
What if we want some per package information?
Well, to do it on one package (let's try rlang):

```{r}
rlang_tally <- tally_up_lofun_c(los$rlang)
rlang_tally$tot_args
rlang_tally$tot_mono_args_type
format_df_for_print(rlang_tally$top_poly_type)
```

As you'll see, the numbers here are low because we didn't actually trace rlang itself, just that some other packages used rlang, giving us limited results.

Now, how about if we take all of R's base classes (i.e., those which occur naturally in R) and lump them all into a "primitive" class.
What sort of class-based polymorphism do we see, now?

```{r}
los_primitive_class <- lapply(los, function(lof) {lapply(lof, class_to_primitive_df)})
```

```{r}
total_mono_clas_prim <- apply_count_to_lopkg_c(los_primitive_class, args_count_all_mono_class)
total_mono_clas_functions_prim <- apply_count_to_lopkg_c(los_primitive_class, fun_count_all_mono_class)
```

Let's get the top poly class now.
```{r}
poly_class_args_prim <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los_primitive_class, extract_poly_class_args))
poly_class_args_prim <- poly_class_args_prim[order(-poly_class_args_prim$count),] # sort by count
poly_class_args_prim$perc <- round(poly_class_args_prim$count/sum(poly_class_args_prim$count) * 100, 2)
```

```{r}
format_df_for_print(poly_class_args_prim[1:20,])
```

The low percentage numbers suggest that no one user-defined pattern contributes too much to the class-based polymorphism.

Alright, let's do a package tally for all packages.
First, let's make all the classes primitive in our previous collapsed list.

```{r}
los_basic_no_null_real_primitive_class <- lapply(los_basic_no_null_real, function(lof) {lapply(lof, class_to_primitive_df)})
```

Note: this takes a long time to run.
```{r}
all_tallies_all_reduced <- lapply(los_basic_no_null_real_primitive_class, tally_up_lofun_c)
```

```{r}
all_tallies_sizes <- lapply(all_tallies_all_reduced, function(x) x$tot_size_of_poly)
```

Ok.
It's certainly not perfect, but one metric I used for figuring out the size of polymorphism is just by summing up the type-, attribute-, and class-based polymorphism of each argument in a function.
This might give us some leads.
The top 20 polymorphic packages by this metric is:

```{r}
head(sort(unlist(all_tallies_sizes), decreasing=TRUE), 20)
```

Ok.
Let's see what we can get out of coinciding type signatures.
With out reduced type system, we might want to know a little bit more about the "real", "character" type signature, which is by far one of the most common signatures.
We hypothesize that these should appear alongside lists (and possibly vectors).
For lists only, we can do this easily:
```{r}
# for list coinciding
coincide_list_index <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_coinciding_types_index_and_list_reduced_TS)

# just real, character
char_real_sigs <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_X_types, types_to_find=c("real", "character"), strict=TRUE)
```

Let's look for vectors, also.
```{r}
coincide_list_vector_index <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_list_or_vector)
```

73% coincide.
Nice.
Dates might be another possibility?
Let's check:

```{r}
sigs_with_real_char <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_X_types, types_to_find=c("real", "character"), strict=TRUE)
```

```{r}
sigs_with_dates <- get_fun_sigs_matching_pred_from_lopkg(los_basic_no_null_real_primitive_class, does_df_have_date_real_char)
```

Hmm... Not as much as we would have liked, just 20% of the leftover (or 5% of the total).
Why don't we just see what's leftover:
```{r}
just_index <- get_fun_sigs_matching_pred_from_lopkg(los, does_df_have_coinciding_types_index_and_NOT_list_or_vector)
```

This isn't working the way I was expecting.

>>>>

---
title: "Paper Results Section"
output: html_notebook
---

This Notebook walks through the results section of the paper.

First, administrivia.
```{r}
source("../code/data_processing.R") # get the functions we need

library(xtable)
```

Load the data.
```{r, echo=FALSE}
los <- list.files("../data/full_sigs", full.names=TRUE)
names <- list.files("../data/full_sigs")
names(los) <- substr(names, 1, nchar(names)-4)
los <- lapply(los, readRDS)
los <- remove_empties(los)
# annoying strange error that I should fix pops up
```

Get the counts.
```{r}
# argument counts
total_arguments <- apply_count_to_lopkg_c(los, args_count_all)
total_mono_type <- apply_count_to_lopkg_c(los, args_count_all_mono_type)
total_mono_attr <- apply_count_to_lopkg_c(los, args_count_all_mono_attr_pattern)
total_mono_clas <- apply_count_to_lopkg_c(los, args_count_all_mono_class)
total_mono_all <- apply_count_to_lopkg_c(los, args_count_all_mono_all)
```

```{r}
tot_arg_df <- data.frame(
  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
  Count=c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments),
  Percentage=round(c(total_mono_all, total_mono_type, total_mono_clas, total_mono_attr, total_arguments) / total_arguments * 100, 2)
)

xtable(tot_arg_df)
```

@@

```{r}
tot_fun_df <- data.frame(
  Type=c("Full Monomorphic", "Monomorphic in Type", "Monomorphic in Class", "Monomorphic in Attribute Pattern", "Total Seen"),
  Count=c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions),
  Percentage=round(c(total_mono_all_functions, total_mono_type_functions, total_mono_clas_functions, total_mono_attr_functions, total_functions) / total_functions * 100, 2)
)

xtable(tot_fun_df)
```

We Have one thing we want to do first.
Namely, let's get rid of vector/X, scalar/X polymorphism.
```{r}
los_vs <- lapply(los, function(lofuns) {lapply(lofuns, combine_scalar_vector_where_appropriate)})
```

Top signatures, with printing.
This is how you get the top sigs.
```{r}
poly_type_args <- reduce_extracted_lopkgs_to_one(apply_extract_to_lopkg(los, extract_poly_type_args))
poly_type_args <- poly_type_args[order(-poly_type_args$count),] # sort by count
poly_type_args$perc <- round(poly_type_args$count/sum(poly_type_args$count) * 100, 2)
```

Print top 10.
```{r}
format_df_for_print(poly_type_args[1:10,]) -> put_me
xtable(put_me)
```

Now, having noted that:
(1) vector/X, scalar/X is common, and
(2) ints and doubles are basically interchangable
while noting that together these make up a huge portion of the polymorphism, we transform.
```{r}

```

### 2019-03-23 paper/main.tex
plots and new data, with vector/scalar dealt with

@@

The first question we will address is: {\it how polymorphic is R?}  To
answer that, we will first look at the number of monomorphic arguments
across all of the packages we analyzed.  The data can be found in
Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}
\AT{TODO: side by side in minipage?}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Argument Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 448K & 85.50 \\ 
  2 & Monomorphic in Type & 471K & 90.28 \\ 
  3 & Monomorphic in Class & 480K & 92.16 \\ 
  4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
  5 & Total Seen & 521K & -- \\ 
   \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in the FGTS.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Function Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 82K & 66.27 \\ 
  2 & Monomorphic in Type & 94K & 75.74 \\ 
  3 & Monomorphic in Class & 97K & 78.45 \\ 
  4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
  5 & Total Seen & 124K & -- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in the FGTS.}
\end{table}

@@

In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
We see that roughly 10\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

As for functions, roughly 25\% are polymorphic in type, or put differently 25\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.

These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Figure~\AT{eyyy}.

By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}.
One way to account for this would be to collapse scalar and vector types into a vector type:
If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector.

>>>>

\AT{TODO: adjust prose to talk about how vector/scalar is dealt with before this point.}

@@

The first question we will address is: {\it how polymorphic is R?}  To
answer that, we will first look at the number of monomorphic arguments
across all of the packages we analyzed.  The data can be found in
Tables~\ref{tab:argcountsFGTS}~and~\ref{tab:funcountsFGTS}.
\AT{TODO: have these tables be generated automatically}
\AT{TODO: put all counts in one table?}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:03:30 2019
\begin{table}[ht]
\label{tab:argcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Argument Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 457K & 87.75 \\ 
  2 & Monomorphic in Type & 485K & 93.08 \\ 
  3 & Monomorphic in Class & 480K & 92.16 \\ 
  4 & Monomorphic in Attribute Pattern & 480K & 91.99 \\ 
  5 & Total Seen & 521K & --- \\ 
     \hline
\end{tabular}
\caption{Account of {\it argument} polymorphism in the FGTS.}
\end{table}

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 18:10:46 2019
\begin{table}[ht]
\label{tab:funcountsFGTS}
\centering
\begin{tabular}{rlrr}
  \hline
 & Function Polymorphism & Count & Percentage \\ 
  \hline
1 & Full Monomorphic & 86K & 69.44 \\ 
  2 & Monomorphic in Type & 100K & 81.01 \\ 
  3 & Monomorphic in Class & 97K & 78.45 \\ 
  4 & Monomorphic in Attribute Pattern & 96K & 77.47 \\ 
  5 & Total Seen & 124K & --- \\ 
   \hline
\end{tabular}
\caption{Account of {\it function} polymorphism in the FGTS.}
\end{table}

@@

In the tables, ``Full Monomorphic'' denotes monomorphism in type, class, and attribute.
We see that roughly 7\% of all argument positions in R are polymorphic in type.
The numbers for attribute and class polymorphism are included for
completeness, but the important number is that of type polymorphism (as our
method of constructing types sometimes factors in class and attribute).
Specifically for classes, often the class and \code{typeof} a value will
line up (refer to Section~\ref{sec:rlang}).

As for functions, roughly 20\% are polymorphic in type, or put differently 20\% have at least one polymorphic argument or a polymorphic return.
The discrepancy between this and the polymorphicity of arguments is explained by the tendency of some functions, particularly those dealing with models, to have {\it many} arguments.

These numbers alone are insufficient, so we ask: {\it of the polymorphic arguments, what are the most common patterns?}
The answer to that question can be found in Table~\ref{tab:toppolyFDTS}.

% latex table generated in R 3.5.1 by xtable 1.8-3 package
% Fri Mar 22 21:27:37 2019
\begin{table}[ht]
\label{tab:toppolyFDTS}
\centering
\begin{tabular}{lrr}
  \hline
 Polymorphic Type Signature & \# Occurrences & \% of Total Polymorphism \\ 
  \hline
  vector/double, vector/integer & 4838 & 13.42 \\ 
  scalar/double, scalar/integer & 3666 & 10.17 \\ 
  NULL, vector/double & 1463 & 4.06 \\ 
  data.frame, vector/double & 1324 & 3.67 \\ 
  NULL, scalar/character & 1237 & 3.43 \\ 
   NULL, scalar/double & 1174 & 3.26 \\ 
   list$<$any$>$, list$<$double$>$ & 1092 & 3.03 \\ 
   list$<$any$>$, list$<$list$>$ & 767 & 2.13 \\ 
   scalar/character, scalar/double & 674 & 1.87 \\ 
   NULL, vector/character & 640 & 1.78 \\ 
     \hline
\end{tabular}
\caption{Top polymorphic argument signatures in the FDTS.}
\end{table}

By and large, the most common pattern is {\tt scalar/X} and {\tt vector/X}, but we deal with this polymorphism separately:
If a function argument is observed to be both a unit-length and non-unit-length vector, then it's reasonable to assume that the argument is intended to be a vector.
